### 2015-05-22 Harris_2015/manuscript-materials/draft.Rmd
more tweaks

For the larger landscapes, I rescaled the four methods' estimates using linear
regression through the origin so that they all had a consistent
interpretation. In each case, I regressed the "true" $\beta$ coefficient for
each species pair against the model's estimate, re-weighting the pairs so that
each landscape contributed equally to the rescaled estimate^[The null model
generated one $Z$-score outlier greater than 1000, which dominated the
regression and squared error analyses. To reduce its influence on these
results, I changed its value to 32.5, which was the value of the next largest
$Z$-score in the null model's results.]. For each estimate of a species pair's
interactions, I used this regression to calculate the squared error associated
with method that produced it.  Finally, I averaged these squared errors for
each combination of species richness, landscape size, statistical method, and 
presence/absence of environmental filtering across all 12 replicates; the mean
squared errors associated with these subsets of the data determined the proportion
of variance explained by each method under different conditions.

@@

***Larger landscapes.***
The accuracy of the four evaluated methods varied substantially, depending
on the parameters that produced the simulated communities (Figure 3).
In general, however, there was a consistent ordering: the Markov network
explained 54% of the variance overall, followed by partial covariances (32%),
sample covariances (21%), and $Z$ scores from the null model (17%).  
The models' accuracies tended to decline when environmental filters were added, 
particularly when the number of species was small and the effects could not be 
diluted among many pairwise interactions.

@@

The results presented above are very promising, as they show that Markov
networks can recover species' pairwise interactions from observational  data,
even when direct interactions are largely overwhelmed by indirect effects
(e.g. Figure 1) or environmental effects (lower panels of Figure 3). For cases
where it is infeasible to fit a Markov network, these results also indicate
that partial covariances---which can be computed straightforwardly by linear
regression---can often provide an accurate approximation.

Apart from the environmental filters, the simulated landscapes presented here
represent the best-case scenario for these methods.  Future research should
thus examine these models' performance characteristics when the "true"
interaction matrices include guild structure or trophic levels, which could
make the $\beta$ coefficients much more difficult to infer (particularly for
linear approximations like the partial covariance approach;
@loh_structure_2013). On the other hand, ecologists may often have prior 
information about the nature of real species' interaction patterns from 
natural history or ecological experiments, which could substantially reduce
the probability and magnitude of error. The rosalia package [@harris_rosalia_2015]
has built-in mechanisms for incorporating this kind of information, if it can 
be expressed as a prior probability distribution or a penalty on the likelihood.

Additionally, it is important to note that, while partial correlations and
Markov networks both prevent us from mistaking marginal associations for
conditional ones, they cannot tell us the underlying biological mechanism.
Real species co-occurrence patterns will depend on a number of factors---especially
in taxa that emigrate in response to other species---and the
$\beta$ coefficients in Markov networks have to reduce this to a single
number. Thus, experiments and natural history knowledge will generally be
required to pin down the exact nature of the interaction (e.g. who outcompetes
whom).

Despite these limitations, the results with environmental filtering seem to
indicate that the method can be very robust.  Additionally, the fact that
Markov networks provide a likelihood function to optimize makes them highly
extensible, even when it is inconvenient to compute the likelihood exactly.
For example, the mistnet software package for joint species distribution modeling
[@harris_generating_2015] can fit *approximate* Markov networks to large
species assemblages (>300 species) while simultaneously modeling each species'
response to the abiotic environment with complex, nonlinear functions. This
sort of approach, which combines multiple ecological processes, could help
ecologists to disentangle different factors behind the co-occurrence patterns
we observe in nature. Numerous other extensions are possible: similar networks
can be fit with continuous variables, count data, or both
[@lee_learning_2012]. There are even methods [@whittam_species_1981;
@tjelmeland_markov_1998] that would allow the coefficient linking two species
in an interaction matrix to vary as a function of the abiotic environment or
of third-party species that could tip the balance between facilitation and
exploitation [@bruno_inclusion_2003]. Fully exploring these possibilities will
require more research into the various available approximations to the log-likelihood
and to its gradient, in order to balance efficiency, accuracy, and
the ability to generate confidence limits for statistical inference.

By providing precise quantitative expectations about the results of species
interactions, Markov networks have the potential for addressing long-standing
ecological questions. For example, Markov networks can provide a precise
answer to the question of how competition affects species' overall prevalence,
which was a major flash point for the null model debates in the 1980’s
[@strong_ecological_1984]. From Equation 1, one can derive the expected
prevalence of a species in the absence of biotic influences ($\frac{1}{1 +
e^{-\alpha}}$). Any significant difference between this value and the observed
prevalence can be attributed to the $\beta$ coefficients linking this species
to its facilitators and competitors (cf Figure 2D).

This paper only scratches the surface of what Markov networks can do for
ecology. This family of models---particularly the Ising model for binary
networks---has been extremely well-studied in statistical physics for nearly a
century, and the models' properties, capabilities, and limits are well-
understood in a huge range of applications, from spatial modeling
[@gelfand_modelling_2005] to neuroscience [@schneidman_weak_2006] to models of
human behavior [@lee_statistical_2013].
Modeling species interactions using the same framework would thus allow
ecologists to tap into an enormous set of existing discoveries and techniques
for dealing with indirect effects, stability, and alternative stable states.

These results also have important implications for the continued use of fixed-fixed
null models in ecology.  The small simulated landscapes described by
Figure 1 show that test statistics based on marginal co-occurrence (such as
C-scores) will not always have a straightforward relationship with the
underlying ecological processes.  Moreover, the larger communities analyzed in
Figure 3 often fell so far outside the null distribution^[Nearly 20% of the
species pairs fell outside the 99.99994% confidence intervals implied by their
$Z$ scores ($|Z| > 5$), and about 10% had uncorrected p-values below R's
default numerical precision of $2 \times 10^{-16}$.] that it probably makes 
more sense to reject whole model rather than to assign blame for the discrepancy to
any one species pair. On average, the pairwise $Z$ scores from the null
model provided less information about direct species interactions than correlation
coefficients did. Researchers using null modeling approaches may
be able to predict twice as much of the variance in species' "true" interaction
strengths using partial covariances from linear regression, or triple them 
using a Markov network.

>>>>

For the larger landscapes, I rescaled the four methods' estimates using linear
regression through the origin so that they all had a consistent
interpretation. In each case, I regressed the "true" $\beta$ coefficient for
each species pair against the model's estimate, re-weighting the pairs so that
each landscape contributed equally to the rescaled estimate. For each 
estimate of a species pair's
interactions, I used this regression to calculate the squared error associated
with method that produced it.  Finally, I averaged these squared errors for
each combination of species richness, landscape size, statistical method, and 
presence/absence of environmental filtering across all 12 replicates; the mean
squared errors associated with these subsets of the data determined the proportion
of variance explained by each method under different conditions.

@@

***Larger landscapes.***
The accuracy of the four evaluated methods varied substantially, depending
on the parameters that produced the simulated communities (Figure 3).
In general, however, there was a consistent ordering: the Markov network
explained 54% of the variance in the "true" parameters, followed by 
partial covariances (33%), and sample covariances (22%).

The null model scores initially explained 12% of the variation. After manually
reducing the value of one especially strong outlier ($Z=1004$, implying
$p<10^{-1000000}$), this increased to 17%.  Figure 3 reflects the adjusted
version of the results.

@@

The results presented above are very promising, as they show that Markov
networks can recover species' pairwise interactions from observational data,
even when direct interactions are largely overwhelmed by indirect effects
(Figure 1) or environmental effects (lower panels of Figure 3). For cases
where fitting a Markov network is infeasible, these results also indicate
that partial covariances---which can be computed straightforwardly by linear
regression---can often provide a surprisingly accurate approximation.
On the other hand, inferences may be much less reliable in cases where the "true"
interaction matrices include guild structure or trophic levels, which could
make the $\beta$ coefficients much more difficult to infer (particularly for
linear methods; @loh_structure_2013). In these cases, it may be necessary to
include prior information about the nature of species' interactions from 
natural history, experiments, or trait data. The rosalia package 
[@harris_rosalia_2015] has built-in mechanisms for incorporating 
this kind of information, if it can be expressed as a prior probability 
distribution on $\beta$ or a penalty on the likelihood.

Ecologists will also need experiments and natural history knowledge to pin
down the exact nature of the interactions identified by the network (e.g. whether
or not negative interactions are consumptive, or which species in a positively-associated
pair is facilitating the other), particularly when real pairs of species can reciprocally 
influence one another's occurrence probabilities in multiple ways simultaneously; the
$\beta$ coefficients in Markov networks have to reduce this complexity to a single
number. In short, partial correlations and Markov networks both prevent us 
from mistaking marginal associations for conditional ones, but they can't tell us the 
underlying biological mechanisms at work.

Despite these limitations, Markov networks have enormous potential to improve
ecological inferences.  For example, Markov networks provide a simple answer
to the question of how competition should affect a species' overall
prevalence, which was a major flash point for the null model debates in the
1980’s [@strong_ecological_1984]. Equation 1 can be used to calculate the
expected prevalence of a species in the absence of biotic influences
[$\frac{1}{1 + e^{-\alpha}}$; @lee_learning_2012]. Competition's effect on
prevalence in a Markov network can thus be calculated by subtracting this value
and the observed prevalence (cf Figure 2D).

Markov networks---particularly the Ising model for binary networks---have been studied 
in statistical physics for nearly a century [@cipra_introduction_1987], 
and the models' properties, 
capabilities, and limits are well-understood in a huge range of applications, 
from spatial statistics [@gelfand_modelling_2005] to neuroscience 
[@schneidman_weak_2006] to models of human behavior [@lee_statistical_2013].
Modeling species interactions using the same framework would thus allow
ecologists to tap into an enormous set of existing discoveries and techniques
for dealing with indirect effects, stability, and alternative stable states
(phase transitions).

This modeling approach is also highly
extensible, even when it is inconvenient to compute the likelihood exactly.
For example, the mistnet software package for joint species distribution modeling
[@harris_generating_2015] can fit *approximate* Markov networks to large
species assemblages (>300 species) while simultaneously modeling each species'
nonlinear response to the abiotic environment. Combining multiple ecological 
processes into a common model could help ecologists to disentangle different 
factors that can confound simpler co-occurrence analyses [cf @connor_checkered_2013]. 
Numerous other extensions are possible: similar networks
can be fit with a mix of discrete and continuous variables, for example
[@lee_learning_2012]. There are even methods [@whittam_species_1981;
@tjelmeland_markov_1998] that would allow the coefficient linking two species
in an interaction matrix to vary as a function of the abiotic environment or
of third-party species that tip the balance between facilitation and
exploitation [@bruno_inclusion_2003].

Finally, the results presented here have important  implications for ecologists' 
continued use of null models to draw inferences about species interactions. The
small simulated data sets described by Figure 1 show that test statistics
based on marginal co-occurrence (such as C-scores) will not always have a
straightforward relationship with the underlying ecological processes. More
generally, deviations from the null model generally provided less information
about direct species interactions than correlation coefficients did.
Scientists currently using null modeling approaches in their research may be
able to capture twice as much of the variance in species' true interaction
strengths using partial covariances from linear regression instead,  or three
times as much using a Markov network.

### 2015-05-22 Harris_2015/manuscript-materials/draft.tex
more tweaks

\documentclass[12pt,]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
    \usepackage{xltxtra,xunicode}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage{setspace}
\usepackage{lineno}
\linenumbers
\usepackage{setspace}
\usepackage{geometry}
\geometry{verbose,letterpaper,tmargin=2.54cm, bmargin=2.54cm,lmargin=2.54cm,rmargin=2.54cm}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={},
            colorlinks=true,
            citecolor=blue,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls
\setlength{\parindent}{0pt}
\setlength{\parskip}{3pt}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\setcounter{secnumdepth}{0}

@@

For the larger landscapes, I rescaled the four methods' estimates using
linear regression through the origin so that they all had a consistent
interpretation. In each case, I regressed the ``true'' \(\beta\)
coefficient for each species pair against the model's estimate,
re-weighting the pairs so that each landscape contributed equally to the
rescaled estimate\footnote{The null model generated one \(Z\)-score
  outlier greater than 1000, which dominated the regression and squared
  error analyses. To reduce its influence on these results, I changed
  its value to 32.5, which was the value of the next largest \(Z\)-score
  in the null model's results.}. For each estimate of a species pair's
interactions, I used this regression to calculate the squared error
associated with method that produced it. Finally, I averaged these
squared errors for each combination of species richness, landscape size,
statistical method, and presence/absence of environmental filtering
across all 12 replicates; the mean squared errors associated with these
subsets of the data determined the proportion of variance explained by
each method under different conditions.

@@

\textbf{\emph{Larger landscapes.}} The accuracy of the four evaluated
methods varied substantially, depending on the parameters that produced
the simulated communities (Figure 3). In general, however, there was a
consistent ordering: the Markov network explained 54\% of the variance
overall, followed by partial covariances (32\%), sample covariances
(21\%), and \(Z\) scores from the null model (17\%).\\The models'
accuracies tended to decline when environmental filters were added,
particularly when the number of species was small and the effects could
not be diluted among many pairwise interactions.

@@

The results presented above are very promising, as they show that Markov
networks can recover species' pairwise interactions from observational
data, even when direct interactions are largely overwhelmed by indirect
effects (e.g.~Figure 1) or environmental effects (lower panels of Figure
3). For cases where it is infeasible to fit a Markov network, these
results also indicate that partial covariances---which can be computed
straightforwardly by linear regression---can often provide an accurate
approximation.

Apart from the environmental filters, the simulated landscapes presented
here represent the best-case scenario for these methods. Future research
should thus examine these models' performance characteristics when the
``true'' interaction matrices include guild structure or trophic levels,
which could make the \(\beta\) coefficients much more difficult to infer
(particularly for linear approximations like the partial covariance
approach; Loh and Wainwright (2013)). On the other hand, ecologists may
often have prior information about the nature of real species'
interaction patterns from natural history or ecological experiments,
which could substantially reduce the probability and magnitude of error.
The rosalia package (Harris 2015a) has built-in mechanisms for
incorporating this kind of information, if it can be expressed as a
prior probability distribution or a penalty on the likelihood.

Additionally, it is important to note that, while partial correlations
and Markov networks both prevent us from mistaking marginal associations
for conditional ones, they cannot tell us the underlying biological
mechanism. Real species co-occurrence patterns will depend on a number
of factors---especially in taxa that emigrate in response to other
species---and the \(\beta\) coefficients in Markov networks have to
reduce this to a single number. Thus, experiments and natural history
knowledge will generally be required to pin down the exact nature of the
interaction (e.g.~who outcompetes whom).

Despite these limitations, the results with environmental filtering seem
to indicate that the method can be very robust. Additionally, the fact
that Markov networks provide a likelihood function to optimize makes
them highly extensible, even when it is inconvenient to compute the
likelihood exactly. For example, the mistnet software package for joint
species distribution modeling (Harris 2015b) can fit \emph{approximate}
Markov networks to large species assemblages (\textgreater{}300 species)
while simultaneously modeling each species' response to the abiotic
environment with complex, nonlinear functions. This sort of approach,
which combines multiple ecological processes, could help ecologists to
disentangle different factors behind the co-occurrence patterns we
observe in nature. Numerous other extensions are possible: similar
networks can be fit with continuous variables, count data, or both (Lee
and Hastie 2012). There are even methods (Whittam and Siegel-Causey
1981, Tjelmeland and Besag 1998) that would allow the coefficient
linking two species in an interaction matrix to vary as a function of
the abiotic environment or of third-party species that could tip the
balance between facilitation and exploitation (Bruno et al. 2003). Fully
exploring these possibilities will require more research into the
various available approximations to the log-likelihood and to its
gradient, in order to balance efficiency, accuracy, and the ability to
generate confidence limits for statistical inference.

By providing precise quantitative expectations about the results of
species interactions, Markov networks have the potential for addressing
long-standing ecological questions. For example, Markov networks can
provide a precise answer to the question of how competition affects
species' overall prevalence, which was a major flash point for the null
model debates in the 1980's (Strong et al. 1984). From Equation 1, one
can derive the expected prevalence of a species in the absence of biotic
influences (\(\frac{1}{1 + e^{-\alpha}}\)). Any significant difference
between this value and the observed prevalence can be attributed to the
\(\beta\) coefficients linking this species to its facilitators and
competitors (cf Figure 2D).

This paper only scratches the surface of what Markov networks can do for
ecology. This family of models---particularly the Ising model for binary
networks---has been extremely well-studied in statistical physics for
nearly a century, and the models' properties, capabilities, and limits
are well- understood in a huge range of applications, from spatial
modeling (Gelfand et al. 2005) to neuroscience (Schneidman et al. 2006)
to models of human behavior (Lee et al. 2013). Modeling species
interactions using the same framework would thus allow ecologists to tap
into an enormous set of existing discoveries and techniques for dealing
with indirect effects, stability, and alternative stable states.

These results also have important implications for the continued use of
fixed-fixed null models in ecology. The small simulated landscapes
described by Figure 1 show that test statistics based on marginal
co-occurrence (such as C-scores) will not always have a straightforward
relationship with the underlying ecological processes. Moreover, the
larger communities analyzed in Figure 3 often fell so far outside the
null distribution\footnote{Nearly 20\% of the species pairs fell outside
  the 99.99994\% confidence intervals implied by their \(Z\) scores
  (\(|Z| > 5\)), and about 10\% had uncorrected p-values below R's
  default numerical precision of \(2 \times 10^{-16}\).} that it
probably makes more sense to reject whole model rather than to assign
blame for the discrepancy to any one species pair. On average, the
pairwise \(Z\) scores from the null model provided less information
about direct species interactions than correlation coefficients did.
Researchers using null modeling approaches may be able to predict twice
as much of the variance in species' ``true'' interaction strengths using
partial covariances from linear regression, or triple them using a
Markov network.

@@

Xiao, X., D. J. McGlinn, and E. P. White. 2015. A strong test of the
Maximum Entropy Theory of Ecology. The American Naturalist 185:E70--E80.

\end{flushleft}
\end{spacing}

>>>>

\documentclass[12pt,]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
    \usepackage{xltxtra,xunicode}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage{setspace}
\usepackage{lineno}
\linenumbers
\usepackage{setspace}
\usepackage{geometry}
\geometry{verbose,letterpaper,tmargin=2.54cm, bmargin=2.54cm,lmargin=2.54cm,rmargin=2.54cm}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={},
            colorlinks=true,
            citecolor=blue,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls
\setlength{\parindent}{0pt}
\setlength{\parskip}{5pt}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\setcounter{secnumdepth}{0}

@@

For the larger landscapes, I rescaled the four methods' estimates using
linear regression through the origin so that they all had a consistent
interpretation. In each case, I regressed the ``true'' \(\beta\)
coefficient for each species pair against the model's estimate,
re-weighting the pairs so that each landscape contributed equally to the
rescaled estimate. For each estimate of a species pair's interactions, I
used this regression to calculate the squared error associated with
method that produced it. Finally, I averaged these squared errors for
each combination of species richness, landscape size, statistical
method, and presence/absence of environmental filtering across all 12
replicates; the mean squared errors associated with these subsets of the
data determined the proportion of variance explained by each method
under different conditions.

@@

\textbf{\emph{Larger landscapes.}} The accuracy of the four evaluated
methods varied substantially, depending on the parameters that produced
the simulated communities (Figure 3). In general, however, there was a
consistent ordering: the Markov network explained 54\% of the variance
in the ``true'' parameters, followed by partial covariances (33\%), and
sample covariances (22\%).

The null model scores initially explained 12\% of the variation. After
manually reducing the value of one especially strong outlier
(\(Z=1004\), implying \(p<10^{-1000000}\)), this increased to 17\%.
Figure 3 reflects the adjusted version of the results.

@@

The results presented above are very promising, as they show that Markov
networks can recover species' pairwise interactions from observational
data, even when direct interactions are largely overwhelmed by indirect
effects (Figure 1) or environmental effects (lower panels of Figure 3).
For cases where fitting a Markov network is infeasible, these results
also indicate that partial covariances---which can be computed
straightforwardly by linear regression---can often provide a
surprisingly accurate approximation. On the other hand, inferences may
be much less reliable in cases where the ``true'' interaction matrices
include guild structure or trophic levels, which could make the
\(\beta\) coefficients much more difficult to infer (particularly for
linear methods; Loh and Wainwright (2013)). In these cases, it may be
necessary to include prior information about the nature of species'
interactions from natural history, experiments, or trait data. The
rosalia package (Harris 2015a) has built-in mechanisms for incorporating
this kind of information, if it can be expressed as a prior probability
distribution on \(\beta\) or a penalty on the likelihood.

Ecologists will also need experiments and natural history knowledge to
pin down the exact nature of the interactions identified by the network
(e.g.~whether or not negative interactions are consumptive, or which
species in a positively-associated pair is facilitating the other),
particularly when real pairs of species can reciprocally influence one
another's occurrence probabilities in multiple ways simultaneously; the
\(\beta\) coefficients in Markov networks have to reduce this complexity
to a single number. In short, partial correlations and Markov networks
both prevent us from mistaking marginal associations for conditional
ones, but they can't tell us the underlying biological mechanisms at
work.

Despite these limitations, Markov networks have enormous potential to
improve ecological inferences. For example, Markov networks provide a
simple answer to the question of how competition should affect a
species' overall prevalence, which was a major flash point for the null
model debates in the 1980's (Strong et al. 1984). Equation 1 can be used
to calculate the expected prevalence of a species in the absence of
biotic influences (\(\frac{1}{1 + e^{-\alpha}}\); Lee and Hastie 2012).
Competition's effect on prevalence in a Markov network can thus be
calculated by subtracting this value and the observed prevalence (cf
Figure 2D).

Markov networks---particularly the Ising model for binary
networks---have been studied in statistical physics for nearly a century
(Cipra 1987), and the models' properties, capabilities, and limits are
well-understood in a huge range of applications, from spatial statistics
(Gelfand et al. 2005) to neuroscience (Schneidman et al. 2006) to models
of human behavior (Lee et al. 2013). Modeling species interactions using
the same framework would thus allow ecologists to tap into an enormous
set of existing discoveries and techniques for dealing with indirect
effects, stability, and alternative stable states (phase transitions).

This modeling approach is also highly extensible, even when it is
inconvenient to compute the likelihood exactly. For example, the mistnet
software package for joint species distribution modeling (Harris 2015b)
can fit \emph{approximate} Markov networks to large species assemblages
(\textgreater{}300 species) while simultaneously modeling each species'
nonlinear response to the abiotic environment. Combining multiple
ecological processes into a common model could help ecologists to
disentangle different factors that can confound simpler co-occurrence
analyses (cf Connor et al. 2013). Numerous other extensions are
possible: similar networks can be fit with a mix of discrete and
continuous variables, for example (Lee and Hastie 2012). There are even
methods (Whittam and Siegel-Causey 1981, Tjelmeland and Besag 1998) that
would allow the coefficient linking two species in an interaction matrix
to vary as a function of the abiotic environment or of third-party
species that tip the balance between facilitation and exploitation
(Bruno et al. 2003).

Finally, the results presented here have important implications for
ecologists' continued use of null models to draw inferences about
species interactions. The small simulated data sets described by Figure
1 show that test statistics based on marginal co-occurrence (such as
C-scores) will not always have a straightforward relationship with the
underlying ecological processes. More generally, deviations from the
null model generally provided less information about direct species
interactions than correlation coefficients did. Scientists currently
using null modeling approaches in their research may be able to capture
twice as much of the variance in species' true interaction strengths
using partial covariances from linear regression instead, or three times
as much using a Markov network.

@@

Xiao, X., D. J. McGlinn, and E. P. White. 2015. A strong test of the
Maximum Entropy Theory of Ecology. The American Naturalist 185:E70--E80.
\end{flushleft}
\end{spacing}

### 2015-05-22 Harris_2015/models.Rmd
more tweaks

As noted in the manuscript, I manually altered one Z-score produced by the Pairs software because
it was such a strong outlier that it completely dominated the regression model described below.

```{r, cache = TRUE}
# The largest predicted Z-score was over 1000; the next largest was about 32
head(sort(pairs_results$predicted, decreasing = TRUE))
pairs_results$predicted[which.max(pairs_results$predicted)] = 32.47
```

@@

```{r, cache = TRUE}

results = do.call(
  rbind,
  lapply(
    list(
      rosalia_results, 
      cov_results, 
      partial_results, 
      pairs_results
    ),
    function(x){
      linear_model = lm(
        truth ~ predicted + 0, 
        data = x, 
        weights = 1/choose(n_spp, 2),
        y = TRUE
      )
      cat(x$method[[1]])
      print(summary(linear_model))
      cat("\n", "\n", "\n")
      residuals = residuals(linear_model)
      cbind(x, residuals = residuals)
    }
  )
)

@@

```{r, cache = TRUE}
library(dplyr)

summarized_results = results %>% group_by_("n_spp", "n_loc", "method", "env") %>% 
  do_(
    ~data.frame(r2 = 1 - sum(.$residuals^2) / sum(.$truth^2))
  )

@@

## Additional assessment

How often does each method get the sign of the interaction correct?
```{r, cache = TRUE}
sapply(
  list(partial = partial_results, pairs = pairs_results, rosalia = rosalia_results, cov = cov_results), 
  function(x){
    signif(100 * mean(sign(x$predicted) == sign(x$truth)), 3)
  }
)
```

Compare with the percent correct one would obtain "by chance" if one knew that 75%
of the true interactions were negative
```{r, cache = TRUE}
signif(
   100 *mean(rbinom(10000, size = 1, prob = .75) == rbinom(10000, size = 1, prob = .75)),
  3
)
```

What percentage of absolute $Z$ scores exceed 5?

```{r}
mean(abs(pairs_results$predicted) > 5)
```

What percentage of absolute $Z$ scores produce p-values below `.Machine$double.eps`?

```{r}
.Machine$double.eps

p_values = pnorm(1 - abs(pairs_results$predicted))
mean(p_values < .Machine$double.eps)
```

>>>>

As noted in the manuscript, the `pairs_results` include one outlier ($Z>1000$) that dominates the 
subsequent analyses.
```{r}
head(sort(pairs_results$predicted, decreasing = TRUE))
```

With the outlier left as-is, the R-squared value is:

```{r}
summary(
  lm(
    truth ~ predicted + 0, 
    data = pairs_results, 
    weights = 1/choose(n_spp, 2),
    y = TRUE
)
)$r.squared
```

For the analyses below, I adjusted the outlier to take on a somewhat more reasonable value.

```{r, cache = TRUE}
pairs_results$predicted[which.max(pairs_results$predicted)] = 32.47 # Second largest result
```

@@

```{r, cache = FALSE}

results = do.call(
  rbind,
  lapply(
    list(
      rosalia_results, 
      cov_results, 
      partial_results, 
      pairs_results
    ),
    function(x){
      linear_model = lm(
        truth ~ predicted + 0, 
        data = x, 
        weights = 1/choose(n_spp, 2),
        y = TRUE
      )
      residuals = residuals(linear_model)
      cbind(x, residuals = residuals)
    }
  )
)

@@

```{r, message=FALSE}
library(dplyr)
```

```{r, cache = FALSE}
summarized_results = results %>% group_by_("n_spp", "n_loc", "method", "env") %>% 
  do_(
    ~data.frame(r2 = 1 - sum(.$residuals^2) / sum(.$truth^2))
  )

@@

```{r}
summarized_results %>% group_by_("method") %>% 
  summarize_(~round(mean(r2) * 100))
```
